---
title: "Conditional branching and looping in R"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
format:
  html:
    theme:
      - cosmo
      - metropolis-page.scss    # same RU look as the single-page presentations
    embed-resources: true
    format-links: false
    include-before-body: page-logo.html
params:
  toMessage: false
---
```{r, setup,include=FALSE, purl=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```

<p>&nbsp;</p>
These exercises are about the conditions and loops sections of [Introduction to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/).

**Exercise 1** - **If Else**

- Build an if statement that test if x is a negative number. Write a print statement that will confirm if x is less than 0. Test it with several values of x.

```{r,echo=toMessage}
x<- 42

if (x<0){
  print("It's a negative number!")
}

x <- -42

if (x<0){
  print("It's a negative number!")
}

```

- Modify the previous if statement to include an else. Ensure a print statement is returned to say if the number is not negative.

```{r,echo=toMessage}

x <- 0

if (x<0){
  print("It's a negative number!")
}else{
  print("It's not a negative number!")
}

x <- -1

if (x<0){
  print("It's a negative number!")
}else{
  print("It's not a negative number!")
}
```

- Finally add a else if statement. We want a response to confirm if x is negative, positive or if it is zero.

```{r,echo=toMessage}
x <- -1

if (x<0){
  print("It's a negative number!")
}else if (x==0){
  print("It's zero")
}else{
  print("It's a positive number!")
}

x <- 0

if (x<0){
  print("It's a negative number!")
}else if (x==0){
  print("It's zero")
}else{
  print("It's a positive number!")
}

x <- 1

if (x<0){
  print("It's a negative number!")
}else if (x==0){
  print("It's zero")
}else{
  print("It's a positive number!")
}

```

- Build an if/else statement that test if a variable is odd/even. Include x in the printed output.

*Hint: The modulus operator may be useful here i.e. x%%2 returns the remainder after the value of x is divided by 2.*

```{r,echo=toMessage}
x <- 1

if (x%%2==0){
  paste(x,"is even")
}else{
  paste(x,"is odd")
}

x <- 2

if (x%%2==0){
  paste(x,"is even")
}else{
  paste(x,"is odd")
}

```

- Using an ifelse() expression, create a factor from a vector of 1 to 40 where all numbers less than 10 are "small",10 to 30 are "mid",31 to 40 are "big"

```{r,echo=toMessage}
condExercise <- 1:40
condExercise
vectorResult <- ifelse(condExercise<10,"small",ifelse(condExercise < 31,"mid","big"))
temp <- factor(vectorResult,levels=c("small","mid","big"),ordered=TRUE)
temp
```

- Calculate the factorial (factorial of 3 = 3 * 2 * 1) of 10 using a loop.

```{r,echo=toMessage}

for(x in 1:10){
  if(x == 1){
    factorialAnswer <- 1
  }else{
    factorialAnswer <- factorialAnswer * x
  }
}
factorialAnswer
```

- Adjusting your answer from before, what is the first number that has a factorial greater than 1000.
```{r,echo=toMessage}

factorialAnswer <- 0
count <- 0

while(factorialAnswer <= 1000){
  count <- count+1
  if(count == 1){
    factorialAnswer <- 1
  }else{
    factorialAnswer <- factorialAnswer * count
  }
}
count
```

- Set your working directory to be in the downloaded course material. Specifically the *docs* subdirectory. [The exact path will depend on where you have saved your download]. Read in all files from the expression directory ("data/ExpressionResults/") with .txt extension and create a table of gene expression results.

```{r,echo=toMessage, eval=FALSE}

setwd("~/Downloads/Intro_To_R_1Day-master/docs")
```

```{r,echo=toMessage}
filesToRead <- dir("data/ExpressionResults/",pattern = "^ExpressionResults_.*\\.txt$",full.names=TRUE)
fileRead <- vector("list",length=length(filesToRead))

for(i in seq_along(filesToRead)){
  fileRead[[i]] <- read.delim(filesToRead[i],header=F,sep="\t")
  colnames(fileRead[[i]]) <- c("GeneNames",basename(filesToRead[i]))
}
mergedTable <- NULL
for(i in fileRead){
  if(is.null(mergedTable)){
    mergedTable <- i
  }else{
    mergedTable <- merge(mergedTable,i,by=1,all=T)
  }

  print(nrow(mergedTable))
}

mergedTable[1:3,]

```

- Add annotation from Annotation.txt. How does the pathway information for genes compare between the expression table and the annotation table?

```{r,echo=toMessage}
Annotation <- read.table("data/ExpressionResults/Annotation.txt",sep="\t",header=TRUE)
annotatedExpression <- merge(Annotation,mergedTable,by=1,all.x=F,all.y=T)
annotatedExpression[1:2,]

table(annotatedExpression$Pathway, useNA = "ifany")
table(Annotation$Pathway, useNA = "ifany")
```
