---
title: "Conditional branching and looping in R answers"
author: "MRC CSC"
date: "01/01/2015"
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,eval = T)
```

These exercises cover the sections of Conditions and Loops in R [Introduction to R]("introToR_Day1.html").

-- Calculate the factorial (factorial of 3 = 3 * 2 * 1) of 10 using a loop.

```{r,message=T}

for(x in 1:10){
  if(x == 1){
    factorialAnswer <- 1
  }else{
    factorialAnswer <- factorialAnswer * x 
  }
}
factorialAnswer
```
-- Adjusting your answer from before, what is the first number that has a factorial greater than 1000.
```{r,message=T}

factorialAnswer <- 0
count <- 0

while(factorialAnswer <= 1000){
  count <- count+1
  if(count == 1){
    factorialAnswer <- 1
  }else{
    factorialAnswer <- factorialAnswer * count 
  }
}
count
```

-- Using an ifelse() expression, create a factor from a vector of 1 to 40 where all numbers less than 10 are "small",10 to 30 are "mid",31 to 40 are "big"

```{r,message=T}
condExercise <- 1:40
condExercise
vectorResult <- ifelse(condExercise<10,"small",ifelse(condExercise < 31,"mid","big"))
temp <- factor(vectorResult,levels=c("small","mid","big"),order=T)
temp
```

-- Read in all files from expression directory with .txt extension and create a table of gene expression results.

```{r,message=T}
exprDirectory <- system.file("extdata", "ExpressionResults/", package = "introToR")

filesToRead <- dir(exprDirectory,pattern = "*\\.txt",full.names=T)
fileRead <- vector("list",length=length(filesToRead))
for(i in 1:length(filesToRead)){
  fileRead[[i]] <- read.delim(filesToRead[i],header=F,sep="\t")
  colnames(fileRead[[i]]) <- c("GeneNames",basename(filesToRead[i]))
}
mergedTable <- NULL
for(i in fileRead){
  if(is.null(mergedTable)){
    mergedTable <- i
  }else{
    mergedTable <- merge(mergedTable,i,by=1,all=T)
  }
  
  print(nrow(mergedTable))
}

mergedTable[1:3,] 

```
-- Add annotation from Annotation.txt. How do the pathway information for genes compare between expression table and annotation table.
 
```{r,message=T}
#Annotation <- read.table("../ExpressionResults/Annotation.ann",sep="\t",h=T)
annFile <- system.file("extdata", "ExpressionResults/Annotation.ann", package = "introToR")

Annotation <- read.table(annFile,sep="\t",h=T)

annotatedExpression <- merge(Annotation,mergedTable,by=1,all.x=F,all.y=T)
annotatedExpression[1:2,]

summary(annotatedExpression$Pathway)
summary(Annotation$Pathway)
```

