---
title: "Reading and writing data in R answers"
author: "MRC CSC"
date: "01/01/2015"
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,eval = T)
```

These exercises cover the  sections of [Introduction to R]("introToR_Day1.html").

**Exercise 1**

Have a look at all files in notepad or excel before attempting to read. All files can be found in the "data" directory.

- Read in the tab delimited file "GeneExpression.txt".
```{r,message=T}

geneExpr <- system.file("extdata", "/data/GeneExpression.txt", package = "reproducibleR")

geneExprMeth <- system.file("extdata", "data/GeneExpressionWithMethods.txt", package = "reproducibleR")

geneExprNotes <- system.file("extdata", "data/GeneExpressionWithNotes.txt", package = "reproducibleR")

#geneExpression <- read.table("../data/GeneExpression.txt",h=T,sep="\t",row.names=1)
geneExpression <- read.table(geneExpr,h=T,sep="\t",row.names=1)
```

- Find the mean expression of all samples.
```{r,message=T}
sampleMeans <- c(mean(geneExpression[,1]),mean(geneExpression[,2]),mean(geneExpression[,3]),mean(geneExpression[,4]),mean(geneExpression[,5]),mean(geneExpression[,5]))
names(sampleMeans) <- colnames(geneExpression)
sampleMeans
# Alternatively we could use the colMeans() function.
colMeans(geneExpression)
```

- Read in the tab delimited file "GeneExpressionWithMethods.txt". This file contains information on analysis steps used to produce file.
```{r,message=T}
#geneExpression <- read.table("../data/GeneExpressionWithMethods.txt",h=T,sep="\t",row.names=1,skip=3)
geneExpression <- read.table(geneExprMeth,h=T,sep="\t",row.names=1,skip=3)
geneExpression
```
- Read in the tab delimited file "GeneExpressionWithNotes.txt". This file contains notes about particular genes.
```{r,message=T}
#geneExpression <- read.table("../data/GeneExpressionWithNotes.txt",h=T,sep="\t",row.names=1,comment.char = ">")
geneExpression <- read.table(geneExprNotes,h=T,sep="\t",row.names=1,comment.char = ">")
geneExpression
```
- Read in tab delimited file "GeneExpression.txt". Order genes by decreasing expression in Sample_1.hi. Write out a new comma separated file with column titles.
```{r,message=T}
#geneExpression <- read.table("../data/GeneExpression.txt",h=T,sep="\t",row.names=1)
geneExpression <- read.table(geneExpr,h=T,sep="\t",row.names=1)
orderedExpression <- geneExpression[order(geneExpression$Sample_1.hi,decreasing=T),]
expressionDF <- cbind(rownames(orderedExpression),orderedExpression)
colnames(expressionDF)[1] <- "geneNames"
write.table(expressionDF,"orderedExpression.txt",sep=",",col.names = T,row.names=F)
```

