---
title: "Conditional branching and looping in R answers"
author: "MRC CSC"
date: "01/01/2015"
output: html_document
---

These exercises cover the sections of Statistics in R [Introduction to R]("introToR_Day1.html").

-- Calculate the factorial (factorial of 3 = 3+2+1) of 10 using a loop.

```{r,message=T}

for(x in 1:10){
  if(x == 1){
    factorialAnswer <- 1
  }else{
    factorialAnswer <- factorialAnswer * x 
  }
}
factorialAnswer
```
-- Adjusting your answer from before, what is the first number that has a factorial greater than 1000.
```{r,message=T}

factorialAnswer <- 0
count <- 0

while(factorialAnswer <= 1000){
  count <- count+1
  if(count == 1){
    factorialAnswer <- 1
  }else{
    factorialAnswer <- factorialAnswer * count 
  }
}
count
```

-- Using an ifelse() expression, create a factor from a vector of 1 to 40 where all numbers less than 10 are "small",10 to 30 are "mid",31 to 40 are "big"

```{r,message=T}
condExercise <- 1:40
condExercise
vectorResult <- ifelse(condExercise<10,"small",ifelse(condExercise < 31,"mid","big"))
temp <- factor(vectorResult,levels=c("small","mid","big"),order=T)
temp
```

-- Read in all files from expression directory with .txt extension and create a table of gene expression results.

```{r,message=T}

filesToRead <- dir("ExpressionResults/",pattern = "*\\.txt",full.names=T)
fileRead <- vector("list",length=length(filesToRead))
for(i in 1:length(filesToRead)){
  fileRead[[i]] <- read.delim(filesToRead[i],header=F,sep="\t")
  colnames(fileRead[[i]]) <- c("GeneNames",basename(filesToRead[i]))
}
mergedTable <- NULL
for(i in fileRead){
  if(is.null(mergedTable)){
    mergedTable <- i
  }else{
    mergedTable <- merge(mergedTable,i,by=1,all=T)
  }
  
  print(nrow(mergedTable))
}

mergedTable[1:3,] 

```
 Add annotation from Annotation.txt. How do the pathway information for genes compare between expression table and annotation table.
 
```{r,message=T}
Annotation <- read.table("ExpressionResults/Annotation.ann",sep="\t",h=T)
annotatedExpression <- merge(Annotation,mergedTable,by=1,all.x=F,all.y=T)
annotatedExpression[1:2,]

table(Annotation$Pathway)
summary(Annotation$Pathway)
```

-- Look at the correlation between samples. Do any samples show a higher degree of correlation.

```{r,message=T}
cor(annotatedExpression[,grep("ExpressionResults",colnames(annotatedExpression))])

```
-- For every gene, perform a t-test and write out the t-statistic (assuming equal variance), log2 fold change, pvalue and gene annotation to a file.
```{r,message=T}
indexGroupOne <- grep("[1-5].txt",colnames(annotatedExpression))
indexGroupTwo <- grep("[6-9,0].txt",colnames(annotatedExpression))
ttestResults <- apply(annotatedExpression,1,function(x)
  t.test(as.numeric(x[indexGroupOne]),as.numeric(x[indexGroupTwo])))

str(ttestResults[[1]])

testResult <- sapply(ttestResults,function(x) c(log2(x$estimate[2]) - log2(x$estimate[1]), x$statistic,x$p.value))
testResult <- t(testResult)
colnames(testResult) <- c("logFC","tStatistic","pValue")
annotatedResult <- cbind(annotatedExpression[,1:3],testResult)
annotatedResult <- annotatedResult[order(annotatedResult$tStatistic),]
annotatedResult[1:2,]
write.table(annotatedResult,file="annotatedResults.csv",sep=",",row.names=F,col.names=F)
```

