---
title: "Introduction to R, Session 2 <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "http://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  xaringan::moon_reader:
    css: ["default", "metropolisCustom.css", "metropolis-fontsCustom.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
  html_document:
    toc: true # table of content true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  isSlides: "no"
---
```{r setup, include=FALSE}
suppressPackageStartupMessages(require(knitr))
knitr::opts_chunk$set(echo = TRUE, tidy = T)
```

```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides != "yes"){
  cat("# Introduction to R (part 2)


"    
  )
  
}

```

## Overview

- [Recap](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session2.html#recap)
- [Conditions and Loops](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session2.html#conditions_and_loops)
- [Libraries](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session2.html#getting_additional_libraries)
- [Writing scripts](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session2.html#scripts)



#  Recap

Session 1 covered introduction to R data types and import/export of data.

- [Background to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session1.html#background-to-r)
- [Data types in R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session1.html#data_types_in_r)
- [Reading and writing in R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session1.html#reading-and-writing-data-in-r)


## Recap

R stores data in five main data types.

- **Vector** - Ordered collection of single data type (numeric/character/logical).
- **Matrix** - Table (ordered 2D collection) of single data type (numeric/character/logical).
- **Factors** - Ordered collection of ordinal or nominal catagories.
- **Data frame** - Table (ordered 2D array) of multiple data types of same length.
- **List** - Ordered collection of multiple data types of differing length.

```{r,echo=T,eval=T}
aVector <- c(1,2,3,4,5,6,7,8,9,10)
aMatrix <- matrix(aVector,ncol=2,nrow=5,byrow = TRUE)
aFactor <- factor(c("R","Python","R","R","Python"),levels = c("R","Python"))
aDataFrame <- data.frame(Number=c(1,2,3,4,5),Factor=aFactor)
```


## Recap

We can access and change information in our data.types using **indexing** with **[]**  (or **[[]]** for lists).

```{r,echo=T,eval=T}
aVector <- c(1,2,3,4,5,6,7,8,9,10)
aVector[10]
aVector[10] <- 0
aVector

```


## Recap
For matricies and data frames we use two indexes, specifying columns and rows. **matrix[_rowIndex_,_columnIndex_]**


```{r,echo=T,eval=T,tidy=FALSE}
aMatrix <- matrix(aVector,ncol=2,
                  nrow=5,byrow = TRUE)
aMatrix
aMatrix[1,]
aMatrix[,1]
```


```{r,echo=T,eval=T,tidy=FALSE}
aMatrix[1,1] <- 0
aMatrix[,2] <- 100
aMatrix
```



## Recap

Remember a matrix can only contain one data type (numeric, character etc). For mixed data types we use data frames.


```{r,echo=T,eval=T,tidy=FALSE}
aMatrix <- matrix(aVector,ncol=2,
                  nrow=5,byrow = TRUE)
aMatrix

```


```{r,echo=T,eval=T,tidy=FALSE}
aMatrix[1,2] <- "Word"
aMatrix
```



## Recap

We can use logical operations to evaluate information our data types.

```{r,echo=T,eval=T}
aVector <- c(1,2,3,4,5,6,7,8,9,10)
aVector > 9
aFactor <- factor(c("R","Python","R","R","Python"),levels = c("R","Python"))
aDataFrame <- data.frame(Number=c(1,2,3,4,5),Factor=aFactor)
aDataFrame
aDataFrame$Factor == "R"
```


## Recap

We can use logical operations with indexing to subset or alter information held in our data types.


```{r,echo=T,eval=T}
aVector <- c(1,2,3,4,5,6,7,8,9,10)
aVector > 5
aVector[aVector > 5] <- 10
aVector
```


## Recap

When replacing factor columns in data frames we have to remember the **levels**.

```{r,echo=T,eval=T,tidy=T}

aFactor <- factor(c("R","Python","R","R","Python"),
                  levels = c("R","Python"))
aDataFrame <- data.frame(Number=c(1,2,3,4,5),
                         Factor=aFactor)
aDataFrame[aDataFrame$Factor == "R",2] <- "NotPython"
aDataFrame
```



## Recap

We can update levels in a data frame column as we would a factor. Now we can replace our elements in data frame.

```{r,echo=T,eval=T,tidy=FALSE}

aDataFrame <- data.frame(Number=c(1,2,3,4,5),
                         Factor=aFactor)
aDataFrame$Factor <- factor(aDataFrame$Factor,
                            levels = c("R","Python","NotPython"))
aDataFrame[aDataFrame$Factor == "R",2] <- "NotPython"
aDataFrame
```



## Recap


Data can be read into R as a table with the **read.table()** function and written to file with the **write.table()** function.

```{r,echo=T,eval=T}
Table <- read.table("data/readThisTable.csv",sep=",",header=T,row.names=1)
Table[1:3,]
```
```{r,echo=T,eval=F}
write.table(Table,file="data/writeThisTable.csv", sep=",", row.names =F,col.names=T)
```



## Recap


If we want to know how a function works or get an example on its use we can use **?** infront function name.

**?_FunctionName_**

```{r,echo=T,eval=F}
?merge
```


```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides == "yes"){
  cat("class: inverse, center, middle

# Conditions and Loops

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 


"    
  )
}else{
  cat("# Conditions and Loops


"    
  )
  
}
```

## Conditions and Loops


We have looked at using logical vectors as a way to index other data types.
```{r}
x <- 1:10
x[x < 4]
```

Logicals are also used in controlling how scripted procedures execute.



## Conditions and Loops

* <span style="color:red">Conditional branching (if,else)</span>
* <span style="color:green">Loops (for, while)</span>

<span style="color:green">While</span> I'm analysing data, <span style="color:red">if</span> I need to execute complex statistical procedures on the data I will use R <span style="color:red">else</span> I will use a calculator.


## Conditional branching


Conditional branching is the evaluation of a logical to determine whether a chunk of code is executed.

In R, we use the **if** statement with the logical to be evaluated in **()** and dependent code to be executed in **{}**.

```{r,message=T}
x <- TRUE
if(x){
  message("x is true")
}
x <- FALSE
if(x){
  message("x is true")
}

```


## Evaluating with if()

More often, we construct the logical value within **()** itself. This can be termed the **condition**. 

```{r,message=T}
x <- 10
y <- 4
if(x > y){
  message("The value of x is ",x," which is greater than ", y)
}
```

The message is printed above because x is greater than y. 

```{r,message=T}
y <- 20
if(x > y){
  message("The value of x is ",x," which is greater than ", y)
}

```
x is now no longer greater than y, so no message is printed.

We really still want a message telling us what was the result of the condition.


## else following an if


If we want to perform an operation when the condition is false we can follow the if() statement with an else statement.

```{r,message=T}
x <- 3
if(x < 5){
  message(x, " is less than to 5")
   }else{
     message(x," is greater than or equal to 5")
}
```


With the addition of the else statement, when x is not less than 5 the code following the else statement is executed.

```{r,message=T}
x <- 10
if(x < 5){
  message(x, " is less than 5")
   }else{
     message(x," is greater than or equal to 5")
}

```




## else if

We may wish to execute different procedures under multiple conditions. This can be controlled in R using the else if() following an initial if() statement.
```{r, message=TRUE}
x <- 5
if(x > 5){
  message(x," is greater than 5")
  }else if(x == 5){
    message(x," is 5")
  }else{
    message(x, " is less than 5")
  }
```


## ifelse()

A useful function to evaluate conditional statements over vectors is the **ifelse()** function.

```{r}
x <- 1:10
x
```

The ifelse() functions take the arguments of the condition to evaluate, the action if the condition is true and the action when condition is false.

```{r, message=TRUE}
ifelse(x <= 3,"lessOrEqual","more") 
```


## ifelse()

We can use multiple nested **ifelse** functions to be apply more complex logical to vectors.

```{r}
ifelse(x == 3,"same",
       ifelse(x < 3,"less","more")
      ) 
```


## Loops

The two main generic methods of looping in R are **while** and **for**

- **while** - *while* loops repeat the execution of code while a condition evaluates as true.

- **for** - *for* loops repeat the execution of code for a range of specified values.


## While loops

While loops are most useful if you know the condition will be satisified but are not sure when (i.e. Looking for a point when a number first occurs in a list).
```{r, message=TRUE}
x <- 1
while(x < 3){
  message("x is ",x," ")
  x <- x+1
}
message("Finally x is not less than 3")
```


## For loops

For loops allow the user to cycle through a range of values applying an operation for every value.

Here we cycle through a numeric vector and print out its value.
```{r, message=TRUE}
x <- 1:5
for(i in x){
  message(i," ", appendLF = F)
}
```

Similarly we can cycle through other vector types (or lists).
```{r, message=TRUE}
x <- toupper(letters[1:5])
for(i in x){
  message(i," ", appendLF = F)
}
```


## Looping through indices

We may wish to keep track of the position in x we are evaluating to retrieve the same index in other variables. A common practice is to loop though all possible index positions of x using the expression **1:length(x)**.

```{r,message=T}
geneName <- c("Ikzf1","Myc","Igll1")
expression <- c(10.4,4.3,6.5)
1:length(geneName)
for(i in 1:length(geneName)){
  message(geneName[i]," has an RPKM of ",expression[i])
}
```


## Loops and conditionals

Loops can be combined with conditional statements to allow for complex control of their execution over R objects. 

```{r,echo=T,eval=F}
x <- 1:13

for(i in 1:13){
  if(i > 10){
    message("Number ",i," is > 10")
  }else if(i == 10){
    message("Number ",i," is  10") 
  }else{
    message("Number ",i," is < 10") 
  }
}
```



```{r,message=T,echo=F}
x <- 1:13

for(i in 1:13){
  if(i > 10){
    message("Number ",i," is greater than 10")
  }else if(i == 10){
    message("Number ",i," is  10") 
  }else{
    message("Number ",i," is less than  10") 
  }
}
```



## Breaking loops

We can use conditionals to exit a loop if a condition is satisfied, just like a while loop.


```{r,message=T,eval=F}
x <- 1:13

for(i in 1:13){
  if(i < 10){
    message("Number ",i," is less than 10")
  }else if(i == 10){
    message("Number ",i," is  10")
    break
  }else{
    message("Number ",i," is greater than  10") 
  }
}
```



```{r,message=T,eval=T,echo=F}
x <- 1:13

for(i in 1:13){
  if(i < 10){
    message("Number ",i,
            " is less than 10")
  }else if(i == 10){
    message("Number ",i,
            " is  10")
    break
  }else{
    message("Number ",i,
            " is greater than  10") 
  }
}
```



## Functions to loop over data types

There are functions which allow you to loop over a data type and apply a function to the subsection of that data.

- **apply** - Apply function to rows or columns of a matrix/data frame and return results as a vector,matrix or list.

- **lapply** - Apply function to every element of a vector or list and return results as a list.

- **sapply** - Apply function to every element of a vector or list and return results as a vector,matrix or list.


## Looping - apply()

The **apply()** function applys a function to the rows or columns of a matrix. The arguments **FUN** specifies the function to apply and **MARGIN** whether to apply the functions by rows/columns or both.

```
apply(DATA,MARGIN,FUN,...)
```

- **DATA** - A matrix (or something to be coerced into a matrix)
- **MARGIN** - 1 for rows, 2 for columns, c(1,2) for cells


## apply() example

```{r}
matExample <- matrix(c(1:4),nrow=2,ncol=2,byrow=T)
matExample
```
Get the mean of rows
```{r}
apply(matExample,1,mean)
```
Get the mean of columns
```{r}
apply(matExample,2,mean)
```


## Additional arguments to apply

Additional arguments to be used by the function in the apply loop can be specified after the function argument. 

Arguments may be ordered as if passed to function directly. For **paste()** function however this isn't possible.


```{r}
apply(matExample,1,paste,collapse=";")
```


## lapply()


Similar to apply, **lapply** applies a function to every element of a vector or list. 

**lapply** returns a list object containing the results of evaluating the function.

```{r}
lapply(c(1,2),mean)
```


As with apply() additional arguments can be supplied after the function name argument.

```{r}
lapply(list(1,c(NA,1),2),mean, na.rm=T)
```


## sapply()


**sapply** (*smart apply*) acts as lapply but attempts to return the results as the most appropriate data type.

Here sapply returns a vector where lapply would return lists.
```{r}
exampleVector <- c(1,2,3,4,5)
exampleList <- list(1,2,3,4,5)
sapply(exampleVector, mean, na.rm=T)
sapply(exampleList, mean, na.rm=T)
```


## sapply() example


In this example lapply returns a list of vectors from the quantile function.

```{r,tidy=FALSE}
exampleList <- list(row1=1:5, 
                    row2=6:10, 
                    row3=11:15)
exampleList
```



```{r}
lapply(exampleList, quantile)
```



## sapply() example 2

Here is an example of sapply parsing a result from the quantile function in a *smart* way.

When a function always returns a vector of the same length, sapply will create a matrix with elements by column.

```{r}
sapply(exampleList, quantile)
```


## sapply() example 3


When sapply cannot parse the result to a vector or matrix, a list will be returned.

```{r,tidy=FALSE}
exampleList <- list(df=data.frame(sample=paste0("patient",1:2), data=c(1,12)),
                    vec=c(1,3,4,5))
sapply(exampleList, summary)
```



## Custom functions
You can create your own functions easily in R using the **<-** assignment. We use the **function** statement with all input variables in **()** and all code to be executed in **{}**. Anything to be received back from the function is indicated by the **return** statement.  

In this case we can recapitulate the square function. 
```{r makingfunctions}
square_my_number  <- function(my_number){
  my_square <- my_number*my_number
  return(my_square)
}

square_my_number(my_number=3)

```


```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides == "yes"){
  cat("class: inverse, center, middle

# Custom functions

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 


"    
  )
}else{
  cat("# Custom functions


"    
  )
  
}
```


## Custom functions
User defined functions can be as long and complicated as you want, taking in many variables, printing out messages and utilizing other functions within them.

Lets make z-scores:
```{r makingfunctions_zscores}
my_zscore  <- function(my_number, my_vector){
  my_mean <- mean(my_vector)
  message("Mean is ", my_mean)
  diff_from_mean <- my_number-my_mean
  stdev <- sd(my_vector)
  my_z <- diff_from_mean/stdev
  return(my_z)
}

A <- rnorm(20)

my_zscore(my_number=A[1], my_vector=A)

```


## Custom functions and apply

These custom functions can also be utilised with apply. 

```{r, message=F}
sapply(A, my_zscore, my_vector=A)

```


## Time for an exercise!


Exercise on loops and conditional branching can be found [here](http://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/exercises//conditionsAndLoops_Exercises.html)


## Answers to exercise.


Answers can be found here  [here](http://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/answers/conditionsAndLoops_Answers.html)

Rcode for answers can be found here  [here](http://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/answers/conditionsAndLoops_Answers.R)


```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides == "yes"){
  cat("class: inverse, center, middle

# Getting Additional Libraries

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 


"    
  )
}else{
  cat("# Getting Additional Libraries


"    
  )
  
}
```


## Loading libraries

Libraries can be loaded using the library() function with an argument of the name of the library.

```{r,eval=F}
library(ggplot2)
```

You can see what libraries are available in the Packages panel or by the library() function with no arguments supplied.

```{r,eval=F}
library()
```


## Installing libraries


Libraries can be installed through the R studio menu.

**-> Tools -> Install packages ..**

Or by using the install.packages() command.

```{r,eval=F}
install.packages("Hmisc")
```


```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides == "yes"){
  cat("class: inverse, center, middle

# Scripts

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 


"    
  )
}else{
  cat("# Scripts


"    
  )
  
}
```



## Saving scripts

Once we have got our functions together and know how we want to analyse our data, we can save our analysis as a **script**. By convention R scripts typically end in **.r** or **.R**

To save a file in RStudio.


**-> File -> Save as**


To open a previous R script

**->File -> Open File..**

To save all the objects (workspace) with extension **.RData**

**->Session -> Save workspace as**


## Sourcing scripts


R scripts allow us to save and reuse custom functions we have written.  To run the code from an R script we can use the **source()** function with the name of the R script as the argument. 

The file **dayOfWeek.r** in the "scripts" directory contains a simple R script to tell you what day it is.

```
#Contents of dayOfWeek.r
dayOfWeek <- function(){
  return(gsub(" .*","",date()))  
}
```
```{r}
source("scripts/dayOfWeek.r")
dayOfWeek()
```


## Rscript


R scripts can be run non-interactively from the command line with the **Rscript** command, usually with the option **--vanilla** to avoid saving or restoring workspaces. All messages/warnings/errors will be output to the console.

```
Rscript --vanilla myscript.r
```

An alternative to Rscript is **R CMD BATCH**. Here all messages/warnings/errors are directed to a file and the processing time appended.

```
R CMD BATCH myscript.r
```


## Sending arguments to Rscript


To provide arguments to an R script at the command line we must add **commandArgs()** function to parse command line arguments.

```{r,eval=F}
args <- commandArgs(TRUE)
myFirstArgument <- args[1]
myFirstArgument
```

```
'10'
```
```{r,eval=F}
as.numeric(myFirstArgument)
```
```
10
```
Since vectors can only be one type, all command line arguments are strings and must be converted to numeric if needed with **as.numeric()**.


## Getting help

- From us: Raise an issue on our [GitHub](https://github.com/RockefellerUniversity/Intro_To_R_1Day/issues). This can be suggestions, comments, edits or questions (about content or the slides themselves).
- Google
- Local friendly bioinformaticians and computational biologists.
- [Stackoverflow](http://stackoverflow.com/)
- [R-help](https://stat.ethz.ch/mailman/listinfo/r-help)

