---
title: "**Introduction to R**"
author: "Bioinformatics Resource Centre"
output:
  html_document:
    toc: true
    toc_float: true
    style: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pkgName="introToR"
name="Intro_To_R_1Day"
repo="RockefellerUniversity"
courseFolder="r_course"
branch="master"
subdir="introToR"

params=list(
  name=name,
  subdir=subdir,
  pkg=pkgName,
  repo=repo,
  branch=branch)
```

# **The Course**

```{r courseOverview, echo=FALSE,eval=TRUE,include=FALSE}
courseSetUp <- system.file("extdata/_course.yml",mustWork = TRUE,package=params$pkg)
courseSetUpYml <- yaml::read_yaml(courseSetUp)
courseDescriptionRMD <- system.file(file.path("extdata","Descriptions",courseSetUpYml$Description$OverviewRmd),mustWork = TRUE,package=params$pkg)
file.copy(courseDescriptionRMD,getwd())
```

```{r showSysInstall, child = basename(courseDescriptionRMD)}
```


# **Setting up**

## System Requirements

```{r installSysReqsMap,echo=FALSE,eval=TRUE,include=FALSE}
tem <- utils::packageDescription(params$pkg)
dependsR <- c("R")
dependsSysReqs <- pkgload::parse_deps(tem$SystemRequirements)$name
SysReqs <- c(dependsR,dependsSysReqs)
listOfSysReqRMDs <- dir(system.file("extdata/SystemRequirements",mustWork = TRUE,package="compileCourses"),full.names = TRUE)
dependsSysReqRMDs <- basename(listOfSysReqRMDs)[basename(listOfSysReqRMDs) %in% paste0("Install_",SysReqs,".Rmd")]
file.copy(listOfSysReqRMDs,dependsSysReqRMDs)
```


```{r showSysInstall, child = dependsSysReqRMDs}
```

## Install required packages.


### From the course package

```{r installFromGithub,echo=FALSE,eval=TRUE,results="asis"}
cat("```{r}\n")
cat("install.packages('BiocManager')\n")
if(params$subdir == ""){
  cat(sprintf("BiocManager::install('%s')",file.path(repo,params$name)),sep = "\n")
}else{
    cat(sprintf("BiocManager::install('%s',subdir='%s')",file.path(params$repo,params$name),params$subdir),sep = "\n")
}
cat("```\n")
```

### From CRAN and Bioconductor
```{r installAllSep,echo=FALSE,eval=TRUE,results="asis"}
tem <- utils::packageDescription(params$pkg)
pkdDep <- unique(c(pkgload::parse_deps(tem$Depends)$name,pkgload::parse_deps(tem$Suggests)$name,pkgload::parse_deps(tem$Imports)$name))
cat("```{r}\n")
cat("install.packages('BiocManager')\n")
cat(sprintf("BiocManager::install('%s')",pkdDep),sep = "\n")
cat("```\n")
```

## Download the material

```{r downloadMaterial,echo=FALSE,eval=TRUE}
repoFull <- file.path("https://github.com",params$repo,params$name,"archive",paste0(params$branch,".zip"))
```

[Download the material](`r repoFull`)

# **The presentations** 

```{r tempa,echo=FALSE,eval=TRUE,include=FALSE}
presTitles <- unlist(strsplit(courseSetUpYml$Presentations$titles," "))
presRMDs <- unlist(strsplit(courseSetUpYml$Presentations$PresRmd," "))
presOverviews <- unlist(strsplit(courseSetUpYml$Presentations$PresOverviewRmd," "))
presRMDsPaths <- file.path("presentations","slides",gsub(".Rmd",".html",basename(presRMDs)))
pageRMDsPaths <- file.path("presentations","singlepage",gsub(".Rmd",".html",basename(presRMDs)))
codeRMDsPaths <- file.path("presentations","r_code",gsub(".Rmd",".R",basename(presRMDs)))
descriptionDir <- system.file(file.path("extdata","Descriptions"),mustWork = TRUE,package=params$pkg)
file.copy(file.path(descriptionDir,presOverviews),presOverviews)
```

```{r temp,echo=FALSE,eval=TRUE,results="asis"}
# titles <- c("A","B")
# descriptions <- c("mdds","asss")
# linksToPres <- c("https://rockefelleruniversity.github.io/Genomic_HeatmapsAndProfiles/r_course/presentations/slides/GenomicHeatmapsAndProfiles2.html#1","https://rockefelleruniversity.github.io/Genomic_HeatmapsAndProfiles/r_course/presentations/slides/GenomicHeatmapsAndProfiles2.html#1")

for(i in 1:length(presRMDs)){
  cat("\n##",presTitles[i],"\n")
  
  # cat("### Description \n")  
   
  # cat(descriptions[i])
  myRes <- knitr::knit_child(presOverviews[i],quiet = TRUE)
  cat(myRes)
  
  cat("\n")  
  cat("\n<iframe src=",presRMDsPaths[i]," width='100%' height='400' title=",presTitles[i],"></iframe>\n",sep="")
  cat("\n") 
  cat("\n") 
  cat("\n### *Links to presentations*\n")
    
  cat("The html slide presentation can be found at this link [**Slide**]","","(",presRMDsPaths[i],")","\n",sep="")
  cat("\n") 
  cat("The single page html presentation can be found at this link [**Single Page**]","","(",pageRMDsPaths[i],")","\n",sep="")
  cat("\n") 
  cat("The code use in the presentations can be found at [**R code**]","","(",codeRMDsPaths[i],")","\n",sep="")
  cat("\n") 
}
```

