---
title: "Introduction to R"
author: "Bionformatics Resource Centre"
date: "5/27/2020"
output: html_document
params:
  pkg: "Intro_To_R_1Day"
  subdir: "introToR"
  repo: "RockefellerUniversity"    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Course

```{r courseDescription,echo=FALSE,eval=TRUE,results="asis"}

```

## Setting up

### System Requirements

### Install required packages.

#### From the course package

```{r installFromGithub,echo=FALSE,eval=TRUE,results="asis"}

cat("```{r}\n")
cat("install.packages('BiocManager')\n")
if(params$subdir == ""){
  cat(sprintf("BiocManager::install('%s')",file.path(repo,params$pkg)))
}else{
    cat(sprintf("BiocManager::install('%s1',subdir=%s2)",file.path(params$repo,params$pkg),params$subdir))
}
cat("```\n")
```

#### From CRAN and Bioconductor
```{r installAllSep,echo=FALSE,eval=TRUE,results="asis"}
tem <- utils::packageDescription(params$pkg)
pkdDep <- unique(c(pkgload::parse_deps(tem$Depends)$name,pkgload::parse_deps(tem$Suggests)$name,pkgload::parse_deps(tem$Imports)$name))
cat("```{r}\n")
cat(sprintf("BiocManager::install('%s')",pkdDep),sep = "\n")
cat("```\n")
```

### Download the material


### The presentations 


