---
title: "Intro"
author: "Binformatics Resource Centre"
date: "5/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Course

This course introduces R and statistical programming.

The course consists of 2 sections, 1 on basic R data types and data input/output and a second on conditionals and looping.
Each section is presented as HTMl presentations or single page document. Exercises and answer sheets are included after all subsections to practice techniques and provide future reference examples.

Course material and exercises are available to view as rendered HTML at https://rockefelleruniversity.github.io/Intro_To_R_1Day/.
All material is available to download under GPL v2 license.

```{r cars,echo=FALSE,eval=TRUE,results="asis"}
tem <- utils::packageDescription("introToR")
pkdDep <- unique(c(pkgload::parse_deps(tem$Depends)$name,pkgload::parse_deps(tem$Suggests)$name,pkgload::parse_deps(tem$Imports)$name))
cat("```{r}\n")
cat(sprintf("BiocManager::Install('%s')",pkdDep),sep = "\n")
cat("```\n")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=TRUE,eval=FALSE}
BiocManager::Install('testthat')
BiocManager::Install('methods')
BiocManager::Install('ggplot2')
BiocManager::Install('rmarkdown')
BiocManager::Install('knitr')
BiocManager::Install('lattice')
BiocManager::Install('yaml')
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
