---
title: "Reading and writing data in R"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```


<p>&nbsp;</p>
These exercises are about the vector sections of [Introduction to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/). 

**Exercise 1**

Have a look at all files in notepad or excel before attempting to read. All files can be found in the "data" directory.

- Read in the tab delimited file "GeneExpression.txt".
```{r,echo=toMessage}
geneExpression <- read.table("data/GeneExpression.txt",h=T,sep="\t",row.names=1)
```

- Find the mean expression of all samples.
```{r,echo=toMessage}
sampleMeans <- c(mean(geneExpression[,1]),mean(geneExpression[,2]),mean(geneExpression[,3]),mean(geneExpression[,4]),mean(geneExpression[,5]),mean(geneExpression[,5]))
names(sampleMeans) <- colnames(geneExpression)
sampleMeans
# Alternatively we could use the colMeans() function.
colMeans(geneExpression)
```

- Read in the tab delimited file "GeneExpressionWithMethods.txt". This file contains information on analysis steps used to produce file.
```{r,echo=toMessage}
geneExpression <- read.table("data/GeneExpressionWithMethods.txt",h=T,sep="\t",row.names=1,skip=3)
geneExpression
```
- Read in the tab delimited file "GeneExpressionWithNotes.txt". This file contains notes about particular genes.
```{r,echo=toMessage}
geneExpression <- read.table("data/GeneExpressionWithNotes.txt",h=T,sep="\t",row.names=1,comment.char = ">")
geneExpression
```
- Read in tab delimited file "GeneExpression.txt". Order genes by decreasing expression in Sample_1.hi. Write out a new comma separated file with column titles.
```{r,echo=toMessage}
geneExpression <- read.table("data/GeneExpression.txt",h=T,sep="\t",row.names=1)
orderedExpression <- geneExpression[order(geneExpression$Sample_1.hi,decreasing=T),]
expressionDF <- cbind(rownames(orderedExpression),orderedExpression)
colnames(expressionDF)[1] <- "geneNames"
write.table(expressionDF,"orderedExpression.txt",sep=",",col.names = T,row.names=F)
```

