---
title: "Reading and writing data in R"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```


<p>&nbsp;</p>
These exercises are about reading and writing data sections of [Introduction to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/). 

**Exercise 1**

Have a look at all files in notepad or excel before attempting to read. All files can be found in the "data" directory.

- Check your current working directory. Set your working directory to be in the downloaded course material. Specifically the *r_course* subdirectory. [The exact path will depend on where you have save your download]

```{r, eval=F}
getwd()
```

```
[1] "/Users/mattpaul"
```

```{r, eval=F}
setwd("~/Downloads/Intro_To_R_1Day-master/r_course")

```


- Read in the tab delimited file "GeneExpression.txt". Check the data type.
```{r,echo=toMessage}
geneExpression <- read.table("data/GeneExpression.txt",h=T,sep="\t",row.names=1)
is(geneExpression)
```

- Coerce the data frame to a matrix
```{r,echo=toMessage}

geneExpression <- as.matrix(geneExpression)
```

- Find the mean expression of all genes.
```{r,echo=toMessage}
sampleMeans <- c(mean(geneExpression[1,]),mean(geneExpression[2,]),mean(geneExpression[3,]),mean(geneExpression[4,]),mean(geneExpression[5,]),mean(geneExpression[6,]),mean(geneExpression[7,]),mean(geneExpression[8,]))
names(sampleMeans) <- rownames(geneExpression)
sampleMeans
# Alternatively we could use the colMeans() function.
# rowMeans(geneExpression)
```

- Coerce the matrix back to a data frame. Add an additional column with extra gene info "Kidney","Adrenal","Liver","Adrenal","Kidney","Liver","Liver","Kidney". Also add the mean expression as a column.

```{R}
geneExpression <- as.data.frame(geneExpression)
geneExpression$tissue_localisation <- factor(c("Kidney","Adrenal","Liver","Adrenal","Kidney","Liver","Liver","Kidney"))
geneExpression$mean_expr <- sampleMeans
geneExpression
```
- Read in tab delimited file "GeneExpression.txt". Order genes by decreasing expression in mean expression. Filter out genes with expression below 5. Write out a new comma separated file with column titles.
```{r,echo=toMessage}

orderedExpression <- geneExpression[order(geneExpression$mean_expr,decreasing=T),]
filteredExpression <- orderedExpression[orderedExpression$mean_expr>5,]
expressionDF <- cbind(rownames(filteredExpression),filteredExpression)
colnames(expressionDF)[1] <- "geneNames"
write.table(expressionDF,"orderedExpression.txt",sep=",",col.names = T,row.names=F)
```

- Read in the tab delimited file "GeneExpressionWithMethods.txt". This file contains information on analysis steps used to produce file.
```{r,echo=toMessage}
geneExpression <- read.table("data/GeneExpressionWithMethods.txt",h=T,sep="\t",row.names=1,skip=3)
geneExpression
```

- Read in the tab delimited file "GeneExpressionWithNotes.txt". This file contains notes about particular genes.
```{r,echo=toMessage}
geneExpression <- read.table("data/GeneExpressionWithNotes.txt",h=T,sep="\t",row.names=1,comment.char = ">")
geneExpression
```



