---
title: "Functions in R"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```


<p>&nbsp;</p>
These exercises are about the functions sections of [Introduction to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/). 

-- Create a function which takes one number and returns the square of that number

```{r,echo=toMessage}
squareOfNumber <- function(number){
  squareNumber <- number^2
  return(squareNumber)
}

squareOfNumber(3)
```

-- Create a function which takes two numbers and returns the mean

```{r,echo=toMessage}
meanOfNumbers <- function(number,number2){
  meanOfNumbersRes <- mean(c(number,number2))
  return(meanOfNumbersRes)
}

meanOfNumbers(1,3)
```

- Create a function which takes two vectors and returns the mean

```{r,echo=toMessage}
meanOfVectors <- function(vector,vector2){
  meanOfVectorsRes <- mean(c(vector,vector2))
  return(meanOfVectorsRes)
}

meanOfVectors(1,3)
```

- Create a function which takes two numbers and returns the two numbers as a vector and the mean, summary and multiple as a data.frame.

```{r,echo=toMessage}
dfAndVecReturn <- function(number,number2){
  input <- c(number,number2)
  df <- data.frame(mean=mean(c(number,number2)),
                   sum=number+number2,
                   multiple=number*number2
                   )
  return(list(input,df))
}

dfAndVecReturn(1,3)
```



-- Create a function which takes one argument and finds the smallest number whose factorial is greater than that argument.

```{r,echo=toMessage}
findSmallestFactorial <- function(x){
    factorialAnswer <- 0
    count <- 0
    while(factorialAnswer <= x){
      count <- count+1
      if(count == 1){
        factorialAnswer <- 1
      }else{
        factorialAnswer <- factorialAnswer * count 
      }
    }
    return(count)
}

findSmallestFactorial(3000)
findSmallestFactorial(10^100)
```


-- Add a if and else statement in your function to only calculate factorial code if argument is a numeric.

```{r,echo=toMessage}
findSmallestFactorial <- function(x){
  if(!is.numeric(x)){
    message("Please provide a numeric argument!")
  }else{
    factorialAnswer <- 0
    count <- 0
    while(factorialAnswer <= x){
      count <- count+1
      if(count == 1){
        factorialAnswer <- 1
      }else{
        factorialAnswer <- factorialAnswer * count 
      }
    }
    return(count)
  }
}

findSmallestFactorial(3000)
findSmallestFactorial("Hello")
```

