---
title: "Lists"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```


<p>&nbsp;</p>
These exercises are about the Lists sections of [Introduction to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/). 

**Exercise 1**

- Create a list containing a character vector, a numeric matrix and a data frame with 2 or more datatypes.

```{r}
firstElement <- c("A","B","C","D","E")
secondElement <- matrix(1:5,nrow=5,ncol=5)
thirdElement <- data.frame(Sample=c("Sample1","Sample2","Sample3","Sample4"), Age=c(25,21,24,25),factor=c("Smoker","Smoker","NonSmoker","Smoker"))

my_list <- list(firstElement, secondElement, thirdElement)
my_list
```
- Rename each entry: "my_vector", "my_matrix", "my_df"

```{r}

names(my_list) <- c("my_vector", "my_matrix", "my_df")
my_list
```

- Access the second column of the dataframe. Try to access it in at least two different ways.

```{r}
my_list$my_df$Age

my_list[[3]][,2]
```

- Add a list into the fourth slot of our list. This list should contain 2 numeric vectors. The resulting list should be 4 long. Check.  

```{r}
fourthElement <- list(1:5,
                      2:6)

my_list <- c(my_list,list(fourthElement ))
length(my_list)
```

- Access the second vector within the sublist in the 4th position.

```{r}
my_list[[4]][[2]]
```

- Create a list with three numeric vectors: c(5,2,9), c(13,12,6), c(1,3,4)

```{r}
my_list <- list(First=c(5,2,9),Second=c(13,12,6),Third=c(1,3,4))

```

- Flatten the list and turn it into a matrix. 

```{r}
flat_list <- unlist(my_list)
listAsMat <- matrix(flat_list,
                    nrow=length(my_list),
                    ncol=3,
                    byrow=T,
                    dimnames=list(names(my_list)))
listAsMat

```