---
title: "Matrices"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```


<p>&nbsp;</p>
These exercises are about the matrices sections of [Introduction to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/). 

**Exercise 1**

- Create a matrix containing information on gene names Gene_1, Gene_2, Gene_3,Gene_4 expression values 1000, 3000, 10000, 12000 and gene lengths 100, 3000, 200, 1000

```{r,echo=toMessage}
geneNames <- c("Gene_1", "Gene_2", "Gene_3","Gene_4")
expression <- c(1000, 3000, 10000, 12000)
geneLengths <- c(100, 3000, 200, 1000)
geneMatrix <- cbind(expression,geneLengths)
rownames(geneMatrix) <- geneNames
geneMatrix
```

- Update the matrix to contain the expression over the gene length for all genes (Length normalised expression) as final column.
<a name="lne"></a>

HINT - We calculated LNE before in [vectors exercises' bonus question](../answers/vector_answers.html#bonus)

```{r,echo=toMessage}
lne <- geneMatrix[,"expression"]/geneMatrix[,"geneLengths"]
geneMatrix <- cbind(geneMatrix,lne)
geneMatrix
```

- Create a smaller matrix containing genes longer than 200
```{r,echo=toMessage}
smallGeneMatrix <- geneMatrix[geneMatrix[,"geneLengths"] > 200,]
smallGeneMatrix
```

- Create a smaller matrix with just expression and lne columns containing genes longer than 200 and expression greater than 300.
```{r,echo=toMessage}
smallGeneMatrix <- geneMatrix[geneMatrix[,"geneLengths"] > 200 & geneMatrix[,"expression"] > 300,c("expression","lne")]
smallGeneMatrix
```

- Add an additional column to the orginal matrix which contains some GO terms associated with these genes: "Positively Regulates Transcription", "Negatively Regulates Transcription", "Positively Regulates Transcription", "Regulates Autophagy"

```{r,echo=toMessage}
GOterms <- c("Positively Regulates Transcription", "Negatively Regulates Transcription", "Positively Regulates Transcription", "Regulates Autophagy")
geneMatrix_GO <- cbind(geneMatrix,GOterms)
geneMatrix_GO
```

- What is the data type? Use the class(),is.numeric() and is.character() functions to investigate. Compare this to the original matrix.

```{r}
class(geneMatrix_GO)
is.numeric(geneMatrix_GO)
is.character(geneMatrix_GO)
class(geneMatrix)
is.numeric(geneMatrix)
is.character(geneMatrix)
```

- Subset the original matrix to only genes that are involved in Transcription. Make sure the result is a numeric matrix. 

```{r}

geneMatrix_transcription <- geneMatrix[grepl("transcription",GOterms, ignore.case = T),]

geneMatrix_transcription
is.numeric(geneMatrix_transcription)
```

**Bonus Question**

<a name="bonus"></a>
- Calculate the sum of expression and length columns for only genes with length > 100.
```{r,echo=toMessage}
expressionSum <- sum(geneMatrix[geneMatrix[,"geneLengths"] > 100,"expression"])
geneLengthSum <- sum(geneMatrix[geneMatrix[,"geneLengths"] > 100,"geneLengths"])
#OR
expressionAndGeneLengthSum <- colSums(geneMatrix[geneMatrix[,"geneLengths"] > 100,c("expression","geneLengths")])
expressionAndGeneLengthSum
```