---
title: "Vector Answers"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```

<p>&nbsp;</p>
These answers are from the exercises about the vector sections of [Introduction to R](https://rockefelleruniversity.github.io/Intro_To_R_1Day/). 



**Exercise 1**

- Create a vector containing the values 1,2,3,4,5
```{r,echo=toMessage}
x <- c(1,2,3,4,5)
x
```
- Create a vector containing the values 1 to 100
```{r,echo=toMessage}
x <- 1:100
x
```
- Create a vector containing the values  0,5,10,15,20
```{r,echo=toMessage}
x <- seq(0,20,5)
x
```
- Create a vector containing the values 1,1,2,2,3,3
```{r,echo=toMessage}
x <- c(rep(1,2),rep(2,2),rep(3,2))
x
```
- Create a vector containing the values 1,1,5,7,9,10
```{r,echo=toMessage}
x <- c(rep(1,2),seq(5,9,2),10)
x
```

**Exercise 2**

- Create a vector containing the values 1 to 10.
```{r,echo=toMessage}
x <- 1:10
x
```
- Create a new vector with all but the first and last value.
```{r,echo=toMessage}
y <- x[-c(1,length(x))]
y
```
- Create a new vector with all but the second and fifth value.
```{r,echo=toMessage}
y <- x[-c(2,5)]
y
```
- Create a new vector of square root of the sixth and seventh position.
```{r,echo=toMessage}
y <- sqrt(x[6:7])
y
```

- Create a new vector of alternating positions in the vector using another vector.
```{r,echo=toMessage}
y <- x[seq(1,10,2)]
y
```


**Exercise 3** 

- Find the number of files in the present working directory.
```{r,echo=toMessage}
# Depend on current working directory..
length(dir())
```

- List the first file in the present working directory.
```{r,echo=toMessage}
x <- dir()
x[1]
```

**Exercise 4**

- Create a vector of the gene names Gene_1, Gene_2, Gene_3
- Create a vector of the expression values 1000, 3000, 10000, 12000
- Create a vector of the gene lengths 100, 3000, 200, 1000
```{r,echo=toMessage}
geneNames <- c("Gene_1", "Gene_2", "Gene_3","Gene_4")
expression <- c(1000, 3000, 10000, 12000)
geneLengths <- c(100, 3000, 200, 1000)
names(expression) <- geneNames
names(geneLengths) <- geneNames
expression
geneLengths
```
- Find the longest gene.
```{r,echo=toMessage}
names(geneLengths[geneLengths == max(geneLengths)])
names(geneLengths[which.max(geneLengths)])
```
- Identify genes which have a length greater than 100 and expression greater than 10000

```{r,echo=toMessage}
geneNames[geneLengths > 100 & expression < 10000]

```

**Bonus Questions**
<a name="bonus" id="bonus"></a>

- Calculate the expression over the gene length for all genes (Length normalised expression).
```{r,echo=toMessage}
lne <- expression/geneLengths
lne
```


- Identify genes with a length normalised expression greater than the average
```{r,echo=toMessage}
geneNames[lne > mean(lne)]

```

 