---
title: "Plotting Excercise"
author: "MRC CSC"
date: "Monday, February 02, 2015"
output: html_document
---

**Exercise 1**
Produce a data frame with two columns: x, which ranges from -2&pi; to 2&pi; 
and has a small interval between values (for plotting), and cosine(x). 
Plot cosine(x) vs. x as a line. 
Repeat, but try some different line types or colours. 
(Hint. you could use seq function)

**Exercise 2**
Produce a plot like the one shown below. 
(Hint: you can use the matplot function to make the plots.) 

```{r,echo=FALSE}
x<-seq(-10,10,0.1) 
y<-sin(x) 
z<-cos(x) 
matplot(x,cbind(y,z),col=c("red","blue"),las=1,xlab="x",ylab="Sine(x) or Cosine(x)",lty=1:2,type="l")
```


**Exercise 3**
Check out the data in the data frame Loblolly (in the datasets package), on the growth of loblolly pine trees from different stocks. 
Create the following plot, or something similar 
(note that you will need to convert height from ft to m).

```{r,echo=FALSE}
ll.dat<-Loblolly 
ll.dat$height<-ll.dat$height*0.3048 
 
plot(height~age,type="n",xlab="Age (yr)",ylab="Height (m)",las=1,xlim=c(0,25),data=ll.dat) 
points(ll.dat$age[ll.dat$Seed==307],ll.dat$height[ll.dat$Seed==329],pch=1,col="red",type="o") 
points(ll.dat$age[ll.dat$Seed==311],ll.dat$height[ll.dat$Seed==307],pch=6,col="blue",type="o") 
points(ll.dat$age[ll.dat$Seed==311],ll.dat$height[ll.dat$Seed==311],pch=22,col="black",type="o") 
legend("topleft",legend=c(311,307,329),pch=c(1,6,22),col=c("red","blue","black"))
```


